Inference of Bifurcations with Differentiable Continuation

### 2021-02-24 docs/article.tex

Alpha justified as being large by balance between numerical instability due to normalisation and desire to constrain the loss to regions where determinant is zero.

@@

But what should the likelihood term \eqref{eq:likelihood} look like if there are no or fewer bifurcations than targets present in the observed region of $p\in\Omega$? This is where the curvature of the determinant $\partial_{\tangent}^2\Det$ can be extremised to encourage bifurcations. We can treat this curvature as an unsupervised correction of magnitude $\lambda$ to a supervised likelihood $L_{\theta}(z|\targets)$. We are now ready to write down the semi-supervised cost function for a bifurcation curve defined by $\rates(z)=0$ in an observed region $p\in\Omega$ as the negative log-likelihood
\begin{align}
    \loss(\theta|\targets,\Omega) := -\log\left(\int_{\partial S}
    L_{\theta}(z|\targets) + \lambda\,K_{\theta}(z)^2\,
    \,\mathrm{d}z\right)\qquad
     K_{\theta}(z):=\frac{\partial^2\Det}{\partial\tangent^2}\\
     \partial S := \{ (u,p) : \rates(u,p)=0, p\in\Omega \}
     \qquad\qquad\qquad\qquad
\end{align}
Both $L_{\theta}(z|\targets)$ and $K_{\theta}(z)$ have explicit analytic expressions and therefore can be differentiated easily. The integration region $\partial S$ is specified implicitly and in most cases will require a numerical procedure to find the roots $\rates(z)=0$ in the region $p\in\Omega$. In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to differentiate the cost $\loss(\theta|\targets,\Omega)$ without having to differentiate through the operations of the solver that finds $\rates(z)=0$. Fortunately this is all possible if we apply a similar strategy to that used by implicit layers \cite{Look2020DifferentiableLayers}; see Appendix \ref{appendix:space-curve} and \ref{appendix:deformation} for derivations.

@@

\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=6cm]{saddle-landscape.png}
\includegraphics[width=6cm]{saddle-optima.png}
\caption{Saddle-node model $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ optimised with respect to targets $\targets=\{-1,1\}$ in observation region $\Omega\in[-2,2]$. Left panel: trajectories in white approaching black line of global optima.}
\label{fig:saddle-node:results}
\end{figure}

>>>>

% Alpha justified as being large by balance between numerical instability due to normalisation and desire to constrain the loss to regions where determinant is zero.

@@

But what should the likelihood term \eqref{eq:likelihood} look like if there are no or fewer bifurcations than targets present in the observed region of $p\in\Omega$? This is where the curvature of the determinant $\partial_{\tangent}^2\Det$ can be extremised to encourage bifurcations. We can treat this curvature as an unsupervised correction of magnitude $\lambda$ to a supervised likelihood $L_{\theta}(z|\targets)$. We are now ready to write down the semi-supervised cost function for a bifurcation curve defined by $\rates(z)=0$ in an observed region $p\in\Omega$ as the negative log-likelihood
\begin{align}
    % \loss(\theta|\targets,\Omega) := \int_{\partial S,\,|\frac{\partial\rates}{\partial u}|=0}
    % \!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\mathrm{d}u\mathrm{d}p\quad\,
    % \frac{1}{|\targets|}\sum_{p'\in\targets}|p-p'|^2
    % -\lambda\log\left(\int_{\partial S}
    % \left(\nabla^2\left|
    % \frac{\partial\rates}{\partial u}
    % \right|\right)^2
    % \,\mathrm{d}u\mathrm{d}p\right)\\
    \loss(\theta|\targets,\Omega) := -\log\left(\int_{\partial S}
    L_{\theta}(z|\targets) + \lambda\,K_{\theta}(z)^2\,
    \,\mathrm{d}z\right)\qquad
     K_{\theta}(z):=\frac{\partial^2\Det}{\partial\tangent^2}\\
     \partial S := \{ (u,p) : \rates(u,p)=0, p\in\Omega \}
     \qquad\qquad\qquad\qquad
\end{align}
Both $L_{\theta}(z|\targets)$ and $K_{\theta}(z)$ have explicit analytic expressions and therefore can be differentiated easily. The integration region $\partial S$ is specified implicitly and in most cases will require a numerical procedure to find the roots $\rates(z)=0$ in the region $p\in\Omega$. In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to differentiate the cost $\loss(\theta|\targets,\Omega)$ without having to differentiate through the operations of the solver that finds $\rates(z)=0$. Fortunately this is all possible if we apply a similar strategy to that used by implicit layers \cite{Look2020DifferentiableLayers}; see Appendix \ref{appendix:space-curve} and \ref{appendix:deformation} for derivations.

@@

\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=7cm]{saddle-landscape.png}
\includegraphics[width=7cm]{saddle-optima.png}
\caption{Saddle-node model $\rates(u,p) = p + \theta_{1}u+\theta_{2}u^3$ optimised with respect to targets $\targets=\{-1,1\}$ in observation region $\Omega\in[-2,2]$. The right panel shows bifurcations diagrams for the three optima marked by stars on the left panel. The optimisation trajectories in white follow the gradient of the cost, approaching the black line of global optima in the left panel}
\label{fig:saddle-node:results}
\end{figure}

\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=7cm]{pitchfork-landscape.png}
\includegraphics[width=7cm]{pitchfork-optima.png}
\caption{Pitchfork model $\rates(u,p) = \theta_{1} + u p +\theta_{2}u^3$ optimised with respect to targets $\targets=\{0\}$ in observation region $\Omega\in[-5,5]$. The right panel shows bifurcations diagrams for the three optima marked by stars on the left panel. The optimisation trajectories in white follow the gradient of the cost, approaching the black line of global optima in the left panel}
\label{fig:pitchfork:results}
\end{figure}
